---
title: "Getting started with risAT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with risAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is risAT?

risAT provides a tidyverse-friendly interface to the Austrian
[RIS](https://www.ris.bka.gv.at/) (Rechtsinformationssystem) Open Government
Data REST API v2.6. It covers the `/Judikatur` endpoint (case law /
jurisprudence) across all supported court applications and is designed for
reproducible legal research.

## Installation

```{r install}
# install.packages("pak")
pak::pak("werkstattcodes/risAT")
```

## Quick search

The easiest way to search is with a court-specific wrapper. Each wrapper
targets a single court application and provides typed, validated arguments:

```{r quick-search}
library(risAT)

# Search the Administrative Court (VwGH) for decisions about "Baurecht"
results <- ris_search_vwgh(query = "Baurecht")
results
```

All search functions return a tibble. Key columns include `id`,
`application`, `decision_date`, `case_number`, `document_type`, and two
list-columns: `content_urls` (download links) and `app_metadata`
(court-specific fields and response metadata).

## Available court wrappers

risAT provides convenience wrappers for the most commonly used Judikatur applications:

| Court / Application | Function |
|---|---|
| VfGH (Constitutional Court) | `ris_search_vfgh()` |
| VwGH (Administrative Court) | `ris_search_vwgh()` |
| Justiz (OGH, OLG, LG, BG) | `ris_search_justiz()` |
| BVwG (Federal Administrative Court) | `ris_search_bvwg()` |
| LVwG (State Administrative Courts) | `ris_search_lvwg()` |
| DSK / DSB (Data Protection) | `ris_search_dsk()` |
| DOK (Disciplinary Bodies) | `ris_search_dok()` |
| PVAK (Staff Representation) | `ris_search_pvak()` |
| GBK (Equal Treatment Commission) | `ris_search_gbk()` |

Plus a generic function that accepts all RIS Judikatur application codes and English aliases documented in `?ris_req_case_law`:

```{r generic-search}
results <- ris_search_case_law(
  application = "data_protection",
  query = "DSGVO"
)
```

## Filtering results

All wrappers accept common filter arguments:

```{r filters}
results <- ris_search_vwgh(
  query = "Asyl",
  decision_date_from = "2024-01-01",
  decision_date_to = "2024-06-30",
  decision_type = "Erkenntnis"
)
```

Some courts have additional parameters. For example, `ris_search_justiz()`
accepts `court`, `legal_area`, and `specialist_area`:

```{r justiz}
results <- ris_search_justiz(
  query = "Schadenersatz",
  court = "OGH"
)
```

## The two-step pattern

Under the hood, each search wrapper calls two lower-level functions:

1. `ris_req_case_law()` -- builds an `httr2` request object
2. `ris_perform_case_law()` -- executes it with automatic pagination

You can use these directly for more control:
```{r two-step}
# Build the request (no network call yet)
req <- ris_req_case_law(
  application = "Vwgh",
  query = "Baurecht"
)

# Inspect the URL before sending
req$url

# Execute with pagination
results <- ris_perform_case_law(req)
```

## Pagination

The RIS API returns results in pages. risAT handles pagination automatically
and fetches all available pages in scope. Pagination metadata for each row is
stored in the `app_metadata` list-column under `response` (API response info)
and `request` (request provenance):

```{r hits}
# Total hit count reported by the API
results$app_metadata[[1]]$response$hits

# Current response page metadata
results$app_metadata[[1]]$response$page_number
results$app_metadata[[1]]$response$page_size

# Request provenance metadata attached by risAT
results$app_metadata[[1]]$request$seitennummer
results$app_metadata[[1]]$request$dokumente_pro_seite
```

## Working with results

The output tibble is designed for use with dplyr and tidyr:

```{r dplyr}
library(dplyr)

results |>
  filter(decision_date >= "2024-01-01") |>
  select(id, case_number, decision_date, decision_type)
```

Content download URLs are stored as a character vector in the `content_urls`
list-column:

```{r content-urls}
# Get URLs for the first result
results$content_urls[[1]]
```

## Verifying results with echo

Whenever relevant and feasible, risAT search functions provide the option to
echo the query to the console. Setting `echo = TRUE` prints the search
arguments and the corresponding URL on the
[RIS website](https://www.ris.bka.gv.at/) that produces the same results.

This serves three purposes:

1. **Verification** -- open the URL in a browser to double-check that the
   results returned by risAT match the official RIS website.
2. **Sharing** -- copy the URL to share a specific query with colleagues who
   may not use R.
3. **Exploration** -- use the RIS website interface to explore the data
   interactively before refining your programmatic query.

```{r echo}
results <- ris_search_vwgh(
  query = "Baurecht",
  decision_date_from = "2024-01-01",
  echo = TRUE
)
```

The printed output includes the total number of hits and a clickable URL
pointing to the equivalent search on `www.ris.bka.gv.at`. The URL preserves
all filter parameters (query, date range, decision type, etc.) so the browser
results correspond exactly to the API results.

All court-specific wrappers and `ris_search_case_law()` support the `echo`
argument.

## Column name reference

The RIS API returns metadata in German. risAT translates these to English
column names. The tables below document all columns grouped by scope, showing
the English column name used by risAT, the original German API field, and a
short description. Columns not listed here are passed through in their
original snake_case form.

### Common columns

These columns appear across most or all court applications.

| Column | German source field | Description |
|---|---|---|
| `id` | `Technisch > ID` | Unique RIS document identifier |
| `application` | `Technisch > Applikation` | RIS application code (e.g. `"Vwgh"`, `"Justiz"`) |
| `authority` | `Technisch > Organ` | Issuing authority / court name |
| `published` | `Allgemein > Veroeffentlicht` | Publication date |
| `modified` | `Allgemein > Geaendert` | Last modification date |
| `document_url` | `Allgemein > DokumentUrl` | RIS web page URL for the document |
| `document_type` | `Dokumenttyp` | Document type (`"Rechtssatz"`, `"Entscheidungstext"`, etc.) |
| `case_number` | `Geschaeftszahl` | Business / case number |
| `norms` | `Normen` | Referenced legal norms (list-column) |
| `decision_date` | `Entscheidungsdatum` | Decision date |
| `keywords` | `Schlagworte` | Keywords / index terms |
| `ecli` | `EuropeanCaseLawIdentifier` | ECLI identifier |
| `full_decision_url` | `GesamteEntscheidungUrl` | URL to the full decision page |
| `legal_principles_url` | `RechtssaetzeUrl` | URL to the legal principles page |
| `decision_text_url` | `EntscheidungstextUrl` | URL to the decision text |
| `content_urls` | `Dokumentliste > ContentReference` | Content download URLs (list-column) |
| `app_metadata` | — | Package-generated metadata (list-column) |

### VwGH (Administrative Court)

| Column | Description |
|---|---|
| `vwgh_decision_type` | Decision type (Beschluss, Erkenntnis, etc.) |
| `vwgh_court` | Court name |
| `vwgh_indices` | Legal index entries (list-column) |
| `vwgh_collection_number` | Official collection number (VwSlg) |
| `vwgh_document_number_type` | Document number type code |
| `vwgh_primary_legal_principle_number` | Primary legal principle number |
| `vwgh_legal_principle_number` | Legal principle number |
| `vwgh_primary_principle_reference` | Reference to the primary legal principle |
| `vwgh_legal_principle_chain_url` | URL to the legal principle chain |
| `vwgh_note` | Editorial notes (Beachte) |
| `vwgh_court_decisions` | Related court decisions (list-column) |

### VfGH (Constitutional Court)

| Column | Description |
|---|---|
| `vfgh_decision_type` | Decision type (Beschluss, Erkenntnis, Vergleich) |
| `vfgh_court` | Court name |
| `vfgh_indices` | Legal index entries (list-column) |
| `vfgh_collection_number` | Official collection number (VfSlg) |
| `vfgh_headnote` | Headnote / guiding principle |
| `vfgh_decision_texts` | Linked decision texts (list-column) |
| `vfgh_decision_text_case_number` | Case number of linked decision text |
| `vfgh_decision_text_document_type` | Document type of linked text |
| `vfgh_decision_text_court` | Court of linked decision text |
| `vfgh_decision_text_decision_date` | Decision date of linked text |
| `vfgh_decision_text_document_url` | URL of linked decision text |
| `vfgh_decision_text_document_number` | Document number of linked text |
| `vfgh_decision_text_decision_type` | Decision type of linked text |

### BVwG (Federal Administrative Court)

| Column | Description |
|---|---|
| `bvwg_decision_type` | Decision type |
| `bvwg_court` | Court name |
| `bvwg_note` | Editorial note |

### LVwG (State Administrative Courts)

| Column | Description |
|---|---|
| `lvwg_decision_type` | Decision type |
| `lvwg_court` | Court name |
| `lvwg_indices` | Legal index entries (list-column) |
| `lvwg_federal_state` | Federal state |
| `lvwg_note` | Editorial note |
| `lvwg_legal_principle_numbers` | Legal principle numbers (list-column) |

### Justiz (Ordinary Courts: OGH, OLG, LG, BG)

| Column | Description |
|---|---|
| `justiz_decision_type` | Decision type |
| `justiz_court` | Court name |
| `justiz_legal_areas` | Legal areas (list-column) |
| `justiz_specialist_areas` | Specialist areas (list-column) |
| `justiz_text_numbers` | Text numbers (list-column) |
| `justiz_legal_principle_numbers` | Legal principle numbers (list-column) |
| `justiz_citation` | Citation / reference |
| `justiz_note` | Editorial note |
| `justiz_decision_texts` | Linked decision texts (list-column) |
| `justiz_decision_text_case_number` | Case number of linked text |
| `justiz_decision_text_document_type` | Document type of linked text |
| `justiz_decision_text_court` | Court of linked text |
| `justiz_decision_text_decision_type` | Decision type of linked text |
| `justiz_decision_text_decision_date` | Decision date of linked text |
| `justiz_decision_text_note` | Note on linked text |
| `justiz_decision_text_document_url` | URL of linked text |

### DSK / DSB (Data Protection Authority)

| Column | Description |
|---|---|
| `dsk_decision_type` | Decision type |
| `dsk_brief_info` | Brief information / summary |
| `dsk_deciding_authority` | Deciding authority |
| `dsk_appeal` | Appeal status |
| `dsk_country` | Country |
| `dsk_language` | Language |
| `dsk_access` | Access level |
| `dsk_decision_on_dsb_document` | Decision on DSB document |
| `dsk_note` | Editorial note |
| `dsk_legal_principle_numbers` | Legal principle numbers (list-column) |

### DOK (Disciplinary Bodies)

| Column | Description |
|---|---|
| `dok_decision_type` | Decision type |
| `dok_brief_info` | Brief information / summary |
| `dok_deciding_authority` | Deciding authority |

### GBK (Equal Treatment Commission)

| Column | Description |
|---|---|
| `gbk_decision_type` | Decision type |
| `gbk_commission` | Commission |
| `gbk_senate` | Senate |
| `gbk_discrimination_ground` | Discrimination ground |
| `gbk_discrimination_offense` | Discrimination offense type |
