---
title: "Getting started with risAT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with risAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.main-container {
  max-width: 650px !important;
}

pre {
  max-height: 600px !important;
  overflow-y: auto !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(risAT)
library(dplyr)
library(ggplot2)
library(lubridate)
```

This vignette walks through the main features of {risAT} using concrete
examples. For a full parameter reference, see the
[function reference](../reference/index.html).

{risAT} provides a tidyverse-friendly interface to the Austrian
[RIS](https://www.ris.bka.gv.at/) (Rechtsinformationssystem) Open Government
Data API v2.6. It covers the `/Judikatur` endpoint — case law from all nine
supported court applications — and the `/Bundesrecht` endpoint for consolidated
federal law, and is designed for reproducible legal research.

# First search

The fastest way in is a court-specific wrapper. Each wrapper targets a single
court and provides typed, validated arguments. Let's retrieve
Administrative Court (VwGH) decisions on housing law (*Baurecht*):

```{r first-search}
library(risAT)

results <- ris_search_vwgh(query = "Baurecht")
results
```

```{r first-search-out, echo=FALSE, eval=TRUE}
results <- readRDS("data/results_vwgh_baurecht.rds")
dplyr::select(results, id, document_type, case_number, decision_date, vwgh_decision_type)
```

All search functions return a tibble — one row per document. The most
frequently used columns are:

| Column | Description |
|---|---|
| `id` | Unique RIS document identifier |
| `case_number` | Business number (*Geschäftszahl*) |
| `decision_date` | Date of the decision |
| `document_type` | `Rechtssatz` or `Text` (full decision) |
| `content_urls` | List-column of download links |
| `app_metadata` | List-column of court-specific and response metadata |

# Filtering results

All wrappers share a common set of filter arguments. Date ranges follow
ISO 8601 (`"YYYY-MM-DD"`), and `decision_type` accepts the German API values
and lowercase aliases.

```{r filters}
results <- ris_search_vwgh(
  query              = "Asyl",
  decision_date_from = "2023-01-01",
  decision_date_to   = "2024-12-31",
  decision_type      = "Erkenntnis"
)

nrow(results)
```

```{r filters-out, echo=FALSE, eval=TRUE}
results_asyl <- readRDS("data/results_vwgh_asyl_filtered.rds")
nrow(results_asyl)
```

The `in_ris_since` argument is a convenient shortcut for recently published
documents:

```{r in-ris-since}
recent <- ris_search_vwgh(in_ris_since = "one_month")
nrow(recent)
```

```{r in-ris-since-out, echo=FALSE, eval=TRUE}
recent <- readRDS("data/results_vwgh_recent.rds")
nrow(recent)
```

# Decisions over time

Because results come back as a tidy tibble, downstream analysis with {dplyr}
and {ggplot2} is straightforward. Below we retrieve VwGH asylum decisions from
2019 to 2024 and visualise the yearly count.

```{r asyl-query}
df_asyl <- ris_search_vwgh(
  query              = "Asyl",
  decision_date_from = "2019-01-01",
  decision_date_to   = "2024-12-31",
  decision_type      = "Erkenntnis"
)
```

```{r asyl-query-out, echo=FALSE, eval=TRUE}
df_asyl <- readRDS("data/results_asyl_plot.rds")
```

```{r asyl-plot, eval=TRUE, fig.width=8, fig.height=4, out.width="100%"}
library(dplyr)
library(ggplot2)
library(lubridate)

df_asyl |>
  mutate(year = year(as.Date(decision_date))) |>
  count(year) |>
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "#2E4057") +
  scale_x_continuous(breaks = 2019:2024) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "VwGH decisions on asylum law by year",
    subtitle = "Search term: 'Asyl', decision type: Erkenntnis, 2019–2024.",
    caption  = "Data: RIS OGD API v2.6; retrieved via {risAT}.",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title          = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle       = element_text(color = "grey40", margin = margin(b = 10)),
    plot.caption        = element_text(color = "grey50", size = rel(0.8)),
    panel.grid.major.x  = element_blank(),
    panel.grid.minor    = element_blank()
  )
```

# Court-specific wrappers

risAT provides wrappers for all nine Judikatur applications:

| Court | Function |
|---|---|
| VfGH (Constitutional Court) | `ris_search_vfgh()` |
| VwGH (Administrative Court) | `ris_search_vwgh()` |
| Justiz (OGH, OLG, LG, BG) | `ris_search_justiz()` |
| BVwG (Federal Administrative Court) | `ris_search_bvwg()` |
| LVwG (State Administrative Courts) | `ris_search_lvwg()` |
| DSK / DSB (Data Protection) | `ris_search_dsk()` |
| DOK (Disciplinary Bodies) | `ris_search_dok()` |
| PVAK (Staff Representation) | `ris_search_pvak()` |
| GBK (Equal Treatment Commission) | `ris_search_gbk()` |

Some wrappers expose court-specific parameters. `ris_search_justiz()` lets you
filter by court and legal area:

```{r justiz}
results_ogh <- ris_search_justiz(
  query      = "Schadenersatz",
  court      = "OGH",
  legal_area = "Zivilrecht"
)
nrow(results_ogh)
```

```{r justiz-out, echo=FALSE, eval=TRUE}
results_ogh <- readRDS("data/results_justiz.rds")
nrow(results_ogh)
```

`ris_search_lvwg()` restricts results to a single state administrative court
via `federal_state`. English state names are accepted:

```{r lvwg}
results_wien <- ris_search_lvwg(
  federal_state = "Vienna",       # also accepted: "Wien"
  decision_type = "Erkenntnis"
)
nrow(results_wien)
```

```{r lvwg-out, echo=FALSE, eval=TRUE}
results_wien <- readRDS("data/results_lvwg.rds")
nrow(results_wien)
```

`ris_search_gbk()` filters Equal Treatment Commission decisions by commission,
senate, and discrimination ground — English aliases are supported:

```{r gbk}
results_gbk <- ris_search_gbk(
  discrimination_ground = "gender",         # "Geschlecht"
  commission            = "private_sector"  # "Gleichbehandlungskommission"
)
nrow(results_gbk)
```

```{r gbk-out, echo=FALSE, eval=TRUE}
results_gbk <- readRDS("data/results_gbk.rds")
nrow(results_gbk)
```

# VfGH: constitutional court decisions

The VfGH wrapper defaults to legal principles (*Rechtssätze*) rather than full
decision texts — the most common use case for constitutional court research.
Toggle via `search_legal_principles` and `search_decision_text`:

```{r vfgh}
# Default: Rechtssätze (RS)
results_vfgh <- ris_search_vfgh(query = "Grundrecht")
nrow(results_vfgh)

# Request full decision texts instead
results_vfgh_et <- ris_search_vfgh(
  query                   = "Grundrecht",
  search_legal_principles = FALSE,
  search_decision_text    = TRUE
)
```

```{r vfgh-out, echo=FALSE, eval=TRUE}
results_vfgh <- readRDS("data/results_vfgh.rds")
nrow(results_vfgh)
```

# Verifying results with echo

Setting `echo = TRUE` prints the search parameters and a direct link to the
same results on the official RIS website. This is useful for:

1. **Verification** — open the URL in a browser and compare with the API results.
2. **Sharing** — paste the URL to share a query with colleagues who don't use R.
3. **Exploration** — browse results on the RIS website before refining your query.

```{r echo}
results <- ris_search_vwgh(
  query              = "Baurecht",
  decision_date_from = "2024-01-01",
  echo               = TRUE
)
```

The printed output shows the total hit count and a clickable RIS URL that
preserves all filter parameters.

# Working with list-columns

Two columns contain richer nested data.

## `content_urls` — download links

Each element of `content_urls` is a character vector of URLs pointing to the
document in different formats (XML, HTML, RTF, PDF):

```{r content-urls}
results$content_urls[[1]]
```

```{r content-urls-out, echo=FALSE, eval=TRUE}
results <- readRDS("data/results_vwgh_baurecht.rds")
results$content_urls[[1]]
```

To extract all PDF URLs across results, filter by file extension:

```{r pdf-urls}
library(purrr)

pdf_urls <- results |>
  mutate(
    pdf_url = map_chr(
      content_urls,
      \(x) {
        hit <- x[grepl("\\.pdf$", x, ignore.case = TRUE)]
        if (length(hit)) hit[[1]] else NA_character_
      }
    )
  ) |>
  filter(!is.na(pdf_url)) |>
  select(case_number, decision_date, pdf_url)

pdf_urls
```

```{r pdf-urls-out, echo=FALSE, eval=TRUE}
library(purrr)

pdf_urls <- results |>
  mutate(
    pdf_url = map_chr(
      content_urls,
      \(x) {
        hit <- x[grepl("\\.pdf$", x, ignore.case = TRUE)]
        if (length(hit)) hit[[1]] else NA_character_
      }
    )
  ) |>
  filter(!is.na(pdf_url)) |>
  select(case_number, decision_date, pdf_url)

pdf_urls
```

## `app_metadata` — court-specific fields and response context

Court-specific fields and pagination context are nested in `app_metadata`.
Use `tidyr::unnest_wider()` to bring them to the top level:

```{r app-metadata}
library(tidyr)

results |>
  select(id, case_number, app_metadata) |>
  unnest_wider(app_metadata) |>
  glimpse()
```

```{r app-metadata-out, echo=FALSE, eval=TRUE}
library(tidyr)

results |>
  select(id, case_number, app_metadata) |>
  unnest_wider(app_metadata) |>
  glimpse()
```

The `response` sub-list contains pagination context (total hits, page number)
and `request` records the exact API parameters sent:

```{r pagination-meta}
# Total hits reported by the API
results$app_metadata[[1]]$response$hits

# Page context
results$app_metadata[[1]]$response$page_number
results$app_metadata[[1]]$response$page_size
```

```{r pagination-meta-out, echo=FALSE, eval=TRUE}
cat(results$app_metadata[[1]]$response$hits, "\n")
cat(results$app_metadata[[1]]$response$page_number, "\n")
cat(results$app_metadata[[1]]$response$page_size, "\n")
```

# Federal law: how often has a law been amended?

Beyond case law, risAT also covers consolidated federal law
(*Bundesrecht in konsolidierter Fassung*) through `ris_search_bundesrecht()`.
Here each row is one consolidated *version* of a provision, and the `amendments`
column records the amending Federal Law Gazettes (*Bundesgesetzblätter*, "BGBl.").
We can use that to answer a common question: **how often has a law been
amended?**

We'll use the *Ökostromgesetz 2012* (Green Electricity Act) as an example:

```{r bundesrecht-search}
oekostrom <- ris_search_bundesrecht(title = "Ökostromgesetz 2012")
nrow(oekostrom)
```

```{r bundesrecht-search-out, echo=FALSE, eval=TRUE}
oekostrom <- readRDS("data/results_bundesrecht_oekostrom.rds")
nrow(oekostrom)
```

The amending acts are listed as free text in the `amendments` column (one
`BGBl.` reference per amendment). Each provision lists the amendments that
affected it, so the *distinct* gazette references across all provisions give the
amendment history of the whole law. We extract them with a regular expression
and deduplicate:

```{r bundesrecht-amend, eval=TRUE}
library(dplyr)
library(stringr)

amendments <- oekostrom |>
  pull(amendments) |>
  str_extract_all("BGBl\\.\\s*[IVX]*\\s*Nr\\.\\s*\\d+/\\d{4}") |>
  unlist() |>
  str_squish() |>
  unique()

# How often has the law been amended?
length(amendments)

sort(amendments)
```

Because each `BGBl.` reference ends in a year, we can also chart the amendment
activity over time:

```{r bundesrecht-amend-plot, eval=TRUE, fig.width=8, fig.height=4, out.width="100%"}
library(ggplot2)

tibble(amendment = amendments) |>
  mutate(year = as.integer(str_extract(amendment, "\\d{4}$"))) |>
  count(year) |>
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "#2E4057") +
  scale_y_continuous(
    breaks = scales::breaks_width(1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = "Amendments to the Ökostromgesetz 2012 by year",
    subtitle = "Distinct amending Federal Law Gazettes (BGBl.) per year.",
    caption  = "Data: RIS OGD API v2.6; retrieved via {risAT}.",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title          = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle       = element_text(color = "grey40", margin = margin(b = 10)),
    plot.caption        = element_text(color = "grey50", size = rel(0.8)),
    panel.grid.major.x  = element_blank(),
    panel.grid.minor    = element_blank()
  )
```

# The two-step pattern

Under the hood every wrapper calls two lower-level functions:

1. `ris_req_case_law()` — builds an `httr2` request object (no network call).
2. `ris_perform_case_law()` — executes the request with automatic pagination.

Use these directly for more control — to inspect the URL before sending, or to
target application codes not covered by a named wrapper:

```{r two-step}
# Build — no network call yet
req <- ris_req_case_law(
  application = "Vwgh",
  query       = "Baurecht"
)

# Inspect the URL before sending
req$url

# Execute — fetches all pages automatically
results <- ris_perform_case_law(req)
nrow(results)
```

```{r two-step-out, echo=FALSE, eval=TRUE}
req <- ris_req_case_law(application = "Vwgh", query = "Baurecht")
cat(req$url, "\n")
```
